# Tutorial 7: Type I & II Errors, Size, and Power

```{r tutorial7-setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  fig.width  = 6.3,
  fig.height = 3.8,
  out.width  = "100%"
)
library(ggplot2)
library(dplyr)
set.seed(42)

theme_tutorial <- function() {
  theme_minimal(base_size = 11) +
    theme(
      plot.title       = element_text(face = "bold", size = 12),
      plot.subtitle    = element_text(color = "grey40", size = 9),
      panel.grid.minor = element_blank(),
      strip.text       = element_text(face = "bold")
    )
}
```

---

## Part 1: Code Walkthrough --- Simulation Guide

This section walks through every simulation step by step. Read it alongside the problem set below. Every simulation is built from scratch.

---

### Two Essential R Functions: `pnorm` and `qnorm`

Before writing any simulation we need to be fluent with two functions. They are the backbone of everything that follows.

#### `pnorm`: the area to the left of a value

`pnorm(x, mean, sd)` answers: *"what fraction of a normal distribution lies below the value x?"*

```{r pnorm-demo}
# The standard normal N(0,1). How much area lies below 1.96?
pnorm(1.96, mean = 0, sd = 1)

# By symmetry, the area above 1.96 is:
1 - pnorm(1.96, mean = 0, sd = 1)

# The area in BOTH tails beyond +/- 1.96 (= alpha for a two-sided test):
(1 - pnorm(1.96)) + pnorm(-1.96)
```

That last line gives us exactly 0.05. This is *why* 1.96 is the magic number for $\alpha = 0.05$.

#### `qnorm`: the inverse --- find the value that gives a target area

`qnorm(p, mean, sd)` answers: *"what value x has exactly fraction p of the distribution below it?"*

```{r qnorm-demo}
# What value cuts off 97.5% of N(0,1) to the left?
qnorm(0.975)    # answer: 1.96 -- the two-sided critical value at alpha = 0.05

# What value cuts off 95% (one-sided critical value)?
qnorm(0.95)     # answer: 1.645

# What value cuts off 99.5% (for alpha = 0.01, two-sided)?
qnorm(0.995)    # answer: 2.576
```

::: {.book-example}
<div class="header">The relationship between the two functions</div>

$\texttt{qnorm}$ and $\texttt{pnorm}$ are inverses of each other:

$$\texttt{pnorm(qnorm(p))} = p \qquad \text{and} \qquad \texttt{qnorm(pnorm(x))} = x$$

The critical value $z_{0.025} = 1.96$ is defined by the equation $\Phi(1.96) = 0.975$. In R: `qnorm(0.975) = 1.96`. We will use this constantly.
:::

```{r verify-inverse}
# Verify: pnorm(qnorm(0.975)) should return exactly 0.975
pnorm(qnorm(0.975))

# Verify: qnorm(pnorm(1.96)) should return exactly 1.96
qnorm(pnorm(1.96))
```

---

### Simulating One Hypothesis Test from Scratch

**Goal:** understand exactly what happens in a single hypothesis test before we repeat it thousands of times.

**Setting:** test $H_0\colon \beta = 0$ vs. $H_1\colon \beta \ne 0$ at $\alpha = 0.05$, with known $\sigma_{\hat{\beta}} = 2$.

#### Step 1 --- Define the world

```{r one-test-step1}
# -- The TRUE state of the world (unknown to the researcher) --
beta_true <- 0    # H0 is TRUE: the real parameter is zero
sigma_hat <- 2    # the known standard deviation of beta_hat

# -- The researcher's decision rule --
alpha  <- 0.05
z_crit <- qnorm(1 - alpha / 2)   # = 1.96 for a two-sided test

cat("True beta:       ", beta_true, "\n")
cat("sigma_beta_hat:  ", sigma_hat, "\n")
cat("Critical value:  ", round(z_crit, 4), "\n")
cat("We reject H0 when |Z| >", round(z_crit, 4), "\n")
```

#### Step 2 --- Observe one estimate

The researcher collects data and computes $\hat{\beta}$. In our simulation, we represent this by drawing one value from $N(\beta_{\text{true}},\, \sigma_{\hat{\beta}}^2)$.

```{r one-test-step2}
set.seed(101)   # fix seed so this specific draw is reproducible

# Draw one OLS estimate from the sampling distribution of beta_hat.
# Under our DGP, beta_hat ~ N(beta_true, sigma_hat^2).
# rnorm(n, mean, sd) draws n values from a normal distribution.
beta_hat_observed <- rnorm(1, mean = beta_true, sd = sigma_hat)

cat("The observed estimate is:  beta_hat =", round(beta_hat_observed, 4), "\n")
cat("The true value is:         beta      =", beta_true, "\n")
cat("Difference (due to noise): ", round(beta_hat_observed - beta_true, 4), "\n")
```

#### Step 3 --- Compute the test statistic

The test statistic standardises the estimate: it measures how many standard deviations $\hat{\beta}$ sits away from the null value $\beta_0 = 0$.

```{r one-test-step3}
# Z = (beta_hat - beta_0) / sigma_hat
# Under H0 (beta_0 = 0): Z = beta_hat / sigma_hat
beta_0 <- 0    # the null hypothesis value

Z_observed <- (beta_hat_observed - beta_0) / sigma_hat

cat("Test statistic:  Z = (", round(beta_hat_observed,3), "- 0) /", sigma_hat,
    "=", round(Z_observed, 4), "\n")
cat("Critical value:  +/-", round(z_crit, 4), "\n")
cat("|Z| =", round(abs(Z_observed), 4), "\n")
```

#### Step 4 --- Make a decision

```{r one-test-step4}
# The rejection rule: reject H0 if |Z| > z_crit
reject_H0 <- abs(Z_observed) > z_crit

if (reject_H0) {
  cat("DECISION: Reject H0  (|Z| =", round(abs(Z_observed),3),
      "> z_crit =", round(z_crit,3), ")\n")
} else {
  cat("DECISION: Fail to reject H0  (|Z| =", round(abs(Z_observed),3),
      "<= z_crit =", round(z_crit,3), ")\n")
}
```

#### Step 5 --- Label the outcome

Since we *know* the true $\beta = 0$ (we set it in Step 1), we can classify this result.

```{r one-test-step5}
# Two possibilities:
#   - If we rejected H0 and H0 is TRUE  -> Type I Error
#   - If we did not reject H0 and H0 is TRUE -> Correct decision

if (reject_H0 & beta_true == 0) {
  cat("OUTCOME: Type I Error (false positive).\n")
  cat("         We rejected H0 even though beta = 0 is TRUE.\n")
} else if (!reject_H0 & beta_true == 0) {
  cat("OUTCOME: Correct. We did not reject H0, and H0 is indeed TRUE.\n")
}
```

::: {.book-example}
<div class="header">What just happened?</div>

We simulated one experiment. The true $\beta = 0$, so $H_0$ is true. Whether we made a Type I error or not depends entirely on the random draw of $\hat{\beta}$. In any one experiment, we cannot tell if we are right or wrong --- we never observe the truth. The 5% guarantee is about what happens *over many repetitions*, which is exactly what we simulate next.
:::

---

### Repeating 10,000 Times: Verifying the Size

**Goal:** repeat the exact procedure from the previous section ten thousand times and count how often we (incorrectly) reject $H_0$ when it is true. The answer should be approximately 5%.

#### Step 1 --- Decide how many repetitions

```{r size-step1}
n_sim     <- 10000   # number of simulated experiments
beta_true <- 0       # H0 is TRUE in all 10,000 experiments
sigma_hat <- 2
z_crit    <- qnorm(0.975)   # = 1.96

cat("We will run", n_sim, "experiments.\n")
cat("In every experiment, the true beta =", beta_true, "(H0 is always TRUE).\n")
cat("We reject whenever |Z| >", round(z_crit, 4), "\n")
```

#### Step 2 --- Draw 10,000 estimates in one line

Instead of using a loop, R lets us draw all 10,000 values at once with a single `rnorm()` call.

```{r size-step2}
# rnorm(n_sim, ...) draws n_sim independent values simultaneously.
# Each value represents beta_hat from one hypothetical experiment.
beta_hat_all <- rnorm(n_sim, mean = beta_true, sd = sigma_hat)

# Inspect the first 10 estimates
cat("First 10 simulated estimates of beta_hat:\n")
round(head(beta_hat_all, 10), 3)
```

```{r size-step2b}
# Verify: the average should be close to the true beta = 0 (unbiasedness)
cat("Mean of 10,000 estimates:", round(mean(beta_hat_all), 4),
    " (should be close to", beta_true, ")\n")
cat("SD of 10,000 estimates:  ", round(sd(beta_hat_all), 4),
    " (should be close to sigma_hat =", sigma_hat, ")\n")
```

#### Step 3 --- Compute all 10,000 test statistics

```{r size-step3}
# Standardise each estimate: Z = beta_hat / sigma_hat
# This produces 10,000 Z-statistics, one per simulated experiment.
Z_all <- beta_hat_all / sigma_hat

cat("First 10 test statistics:\n")
round(head(Z_all, 10), 3)

# Under H0, each Z should look like a draw from N(0,1).
# Check: mean should be ~0, SD should be ~1.
cat("\nMean of Z statistics:", round(mean(Z_all), 4), " (expected: 0)\n")
cat("SD of Z statistics:  ", round(sd(Z_all), 4),   " (expected: 1)\n")
```

#### Step 4 --- Apply the rejection rule to all experiments

```{r size-step4}
# This produces a logical (TRUE/FALSE) vector of length 10,000.
# TRUE  = we rejected H0 in that experiment.
# FALSE = we did not reject H0 in that experiment.
rejected_all <- abs(Z_all) > z_crit

# How many rejections in total?
cat("Total rejections (out of", n_sim, "):", sum(rejected_all), "\n")

# The proportion of rejections = empirical Type I error rate.
# mean() on a TRUE/FALSE vector returns the fraction of TRUEs.
empirical_size <- mean(rejected_all)
cat("Empirical Type I error rate:", round(empirical_size, 4),
    " -- target: alpha =", alpha, "\n")
```

#### Step 5 --- Visualise the distribution of Z statistics

```{r size-step5, fig.height=3.8}
df_size <- data.frame(Z = Z_all, rejected = rejected_all)

ggplot(df_size, aes(x = Z, fill = rejected)) +
  geom_histogram(bins = 80, color = "white", linewidth = 0.1) +
  scale_fill_manual(
    values = c("TRUE" = "#cc2222", "FALSE" = "#aabbd4"),
    labels = c("TRUE"  = paste0("Rejected H0 (Type I error, n = ", sum(rejected_all), ")"),
               "FALSE" = "Correctly not rejected"),
    name = NULL) +
  geom_vline(xintercept = c(-z_crit, z_crit), linetype = "dashed", linewidth = 0.8) +
  annotate("text", x =  z_crit + 0.15, y = 350, label = "+1.96", hjust = 0, size = 3) +
  annotate("text", x = -z_crit - 0.15, y = 350, label = "-1.96", hjust = 1, size = 3) +
  labs(
    title    = paste0("Distribution of Z across ", n_sim, " experiments (H0 true)"),
    subtitle = paste0("Red = Type I errors (", round(empirical_size*100,1),
                      "% of all experiments) -- very close to alpha = 5%"),
    x = "Test statistic  Z = beta_hat / sigma_hat",
    y = "Count"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", plot.title = element_text(face = "bold"))
```

::: {.book-example}
<div class="header">What this histogram tells us</div>

Under $H_0$, the test statistics follow $N(0,1)$ --- the histogram should look like a bell curve centred at zero. The dashed lines at $\pm 1.96$ define the rejection region. The **red** bars (roughly 5% of all bars) are the Type I errors: experiments where $|Z| > 1.96$ purely by chance, even though the truth is $\beta = 0$.

The critical value 1.96 was engineered so that exactly 5% of $N(0,1)$ falls in the tails. This simulation confirms it empirically.
:::

---

### The Power Function: What Happens When H0 is False?

**Goal:** understand how the probability of *correctly* rejecting $H_0$ (= power) depends on the true effect size $\beta^*$.

#### Step 1 --- What changes when H0 is false?

When the true $\beta = \beta^*$ (not zero), the estimate $\hat{\beta}$ is no longer centred at zero --- it is centred at $\beta^*$. The test statistic therefore follows $N(\delta, 1)$ instead of $N(0,1)$, where $\delta = \beta^*/\sigma_{\hat{\beta}}$ is the signal-to-noise ratio.

```{r power-step1}
sigma_hat <- 2
z_crit    <- qnorm(0.975)

# When beta* = 6: the test statistic Z = beta_hat / 2 ~ N(6/2, 1) = N(3, 1)
beta_star  <- 6
delta      <- beta_star / sigma_hat   # signal-to-noise ratio

cat("True effect:          beta* =", beta_star, "\n")
cat("Signal-to-noise:      delta = beta*/sigma_hat =", delta, "\n")
cat("Distribution of Z:    N(delta, 1) = N(", delta, ", 1)\n")
cat("Distribution under H0: N(0, 1)\n")
cat("The distribution has shifted RIGHT by", delta, "units.\n")
```

#### Step 2 --- Compute power for beta* = 6 analytically

Power = $P(|Z| > 1.96)$ when $Z \sim N(\delta, 1)$. We split into two tails.

```{r power-step2}
# RIGHT tail: P(Z > 1.96) when Z ~ N(delta, 1)
# Standardise: let W = Z - delta ~ N(0,1), then Z > 1.96 <=> W > 1.96 - delta
right_tail <- 1 - pnorm(z_crit - delta)   # = P(W > z_crit - delta)

# LEFT tail: P(Z < -1.96) when Z ~ N(delta, 1)
# Z < -1.96 <=> W < -1.96 - delta
left_tail  <- pnorm(-z_crit - delta)       # = P(W < -z_crit - delta)

power_beta6 <- right_tail + left_tail

cat("Right tail contribution: P(Z > 1.96) =", round(right_tail, 4), "\n")
cat("Left  tail contribution: P(Z < -1.96) =", round(left_tail, 4), "\n")
cat("Power at beta* = 6:      ", round(power_beta6, 4),
    "-- i.e.", round(power_beta6*100, 1), "%\n")
```

#### Step 3 --- Verify with simulation

```{r power-step3}
set.seed(42)
n_sim_power <- 10000

# Simulate 10,000 experiments where the TRUE beta = 6 (H0 is FALSE)
beta_hat_h1 <- rnorm(n_sim_power, mean = beta_star, sd = sigma_hat)
Z_h1        <- beta_hat_h1 / sigma_hat    # test statistic (computed as if H0 were true)
rejected_h1 <- abs(Z_h1) > z_crit        # apply the decision rule

simulated_power <- mean(rejected_h1)   # fraction of correct rejections = power

cat("Analytical power at beta* =", beta_star, ":", round(power_beta6, 4), "\n")
cat("Simulated  power at beta* =", beta_star, ":", round(simulated_power, 4), "\n")
cat("They agree: the formula is correct.\n")
```

#### Step 4 --- Compute power for every possible true value

Now we repeat Steps 2--3 for a grid of values $\beta^* \in [-8, 8]$ to draw the full power curve.

```{r power-step4, fig.height=4.0}
# Grid of candidate true values
beta_grid <- seq(-8, 8, by = 0.1)

# For each value on the grid, compute delta and then power analytically
delta_grid     <- beta_grid / sigma_hat
power_curve    <- pnorm(-z_crit - delta_grid) + (1 - pnorm(z_crit - delta_grid))

# Simulation check at a coarser grid (5,000 draws per point)
beta_check  <- seq(-8, 8, by = 2)
power_check <- sapply(beta_check, function(b) {
  Z_check <- rnorm(5000, mean = b, sd = sigma_hat) / sigma_hat
  mean(abs(Z_check) > z_crit)
})

# Plot
df_curve <- data.frame(beta = beta_grid, power = power_curve)
df_check <- data.frame(beta = beta_check, power = power_check)

ggplot() +
  geom_line(data = df_curve, aes(x = beta, y = power),
            color = "#2255aa", linewidth = 1.1) +
  geom_point(data = df_check, aes(x = beta, y = power),
             color = "#cc5500", size = 2.5) +
  geom_hline(yintercept = 0.05, linetype = "dotted", color = "grey50") +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "#2a7a2a") +
  geom_vline(xintercept = 0, color = "grey30", linewidth = 0.4) +
  annotate("text", x = 7.5, y = 0.08, label = "alpha = 0.05 (minimum power)",
           size = 2.8, hjust = 1, color = "grey40") +
  annotate("text", x = 7.5, y = 0.84, label = "80% -- conventional minimum",
           size = 2.8, hjust = 1, color = "#2a7a2a") +
  scale_y_continuous(labels = scales::percent, breaks = seq(0, 1, 0.2)) +
  labs(title    = "Power function: P(reject H0 | true beta = beta*)",
       subtitle = "Blue line = analytical formula;  orange dots = simulation verification",
       x = "True parameter value  beta*",  y = "Power") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

#### Step 5 --- How does sample size move the curve?

```{r power-step5, fig.height=4.0}
# The SE formula from Wooldridge Theorem 2.2:
#   SE(beta_hat) = sigma / sqrt(SST_x)
#
# For x ~ Uniform(0, 10):
#   E[SST_x] = n * Var(x) = n * 10^2/12
#   => SE = sigma_u / sqrt(n * 100/12)
#
# Larger n => smaller SE => larger delta = beta*/SE => higher power

sigma_u <- 4    # error standard deviation
x_range <- 10   # x ~ Uniform(0, x_range)

# For each sample size, recompute the power curve with the appropriate SE
n_values   <- c(20, 50, 200)
power_by_n <- do.call(rbind, lapply(n_values, function(n) {
  se_n      <- sigma_u / sqrt(n * x_range^2 / 12)   # SE for this n
  delta_n   <- beta_grid / se_n
  power_n   <- pnorm(-z_crit - delta_n) + (1 - pnorm(z_crit - delta_n))

  cat("n =", sprintf("%3d", n),
      "  SE(beta_hat) =", round(se_n, 3),
      "  Power at beta*=2:", round(power_n[beta_grid == 2], 3), "\n")

  data.frame(beta = beta_grid, power = power_n, n = as.character(n))
}))

ggplot(power_by_n, aes(x = beta, y = power, color = n)) +
  geom_line(linewidth = 1.1) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("20"="#cc5500","50"="#2255aa","200"="#2a7a2a"),
                     name = "n") +
  scale_y_continuous(labels = scales::percent, breaks = seq(0, 1, 0.2)) +
  labs(title    = "Larger samples shift the entire power curve up",
       subtitle = "Same DGP, same alpha, different n -- only SE changes",
       x = "True beta*", y = "Power") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"), legend.position = "top")
```

---

### The OLS Estimator: Building Sampling Intuition

**Goal:** show concretely that $\hat{\beta}_1^{\text{OLS}}$ is a random variable whose distribution we can characterise.

#### Step 1 --- Define the data-generating process (DGP)

```{r ols-step1}
# These are population-level parameters -- the researcher NEVER observes them.
beta0_true <- 1    # true intercept
beta1_true <- 3    # true slope  (this is what we want to estimate)
sigma_u    <- 4    # standard deviation of the error term u
n_obs      <- 30   # observations per sample

cat("DGP:  y_i = ", beta0_true, "+ ", beta1_true, "* x_i + u_i\n")
cat("      u_i ~ N(0,", sigma_u^2, ")\n")
cat("      x_i ~ Uniform(0, 10)\n")
cat("      n   =", n_obs, "observations per sample\n")
```

#### Step 2 --- Draw one sample and inspect it

```{r ols-step2}
set.seed(7)

# Draw x from Uniform(0, 10) -- satisfies SLR.2 and SLR.3
x_sample <- runif(n_obs, min = 0, max = 10)

# Generate the error term: u ~ N(0, sigma_u^2)
u_sample <- rnorm(n_obs, mean = 0, sd = sigma_u)

# Apply the DGP to get y
y_sample <- beta0_true + beta1_true * x_sample + u_sample

# Look at the first 5 observations
df_sample <- data.frame(x = x_sample, u = u_sample, y = y_sample)
cat("First 5 observations:\n")
print(round(head(df_sample, 5), 3))

cat("\nNote: y = 1 + 3*x + u. For observation 1:\n")
cat("  y =", round(y_sample[1],3), "=",
    beta0_true, "+", beta1_true, "*", round(x_sample[1],3),
    "+", round(u_sample[1],3), "\n")
```

#### Step 3 --- Fit OLS and read off the estimate

```{r ols-step3}
# lm(y ~ x) minimises sum of (y_i - b0 - b1*x_i)^2 over b0 and b1.
# It returns the OLS estimates beta_hat_0 and beta_hat_1.
model      <- lm(y_sample ~ x_sample)
beta0_hat  <- coef(model)[1]   # estimated intercept
beta1_hat  <- coef(model)[2]   # estimated slope

cat("OLS results from this one sample:\n")
cat("  beta_hat_0 (intercept) =", round(beta0_hat, 3),
    "  [true value:", beta0_true, "]\n")
cat("  beta_hat_1 (slope)     =", round(beta1_hat, 3),
    "  [true value:", beta1_true, "]\n")
cat("  Estimation error:       ", round(beta1_hat - beta1_true, 3), "\n")
```

```{r ols-step3b, fig.height=3.8}
ggplot(df_sample, aes(x = x, y = y)) +
  geom_point(color = "#2255aa", size = 2, alpha = 0.8) +
  geom_abline(slope = beta1_true, intercept = beta0_true,
              color = "black", linewidth = 1, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, color = "#cc5500", linewidth = 1.1) +
  annotate("text", x = 0.3, y = max(y_sample) * 0.97,
           label = "True line (dashed)", hjust = 0, size = 3) +
  annotate("text", x = 0.3, y = max(y_sample) * 0.87,
           label = paste0("OLS fit  beta_hat_1 = ", round(beta1_hat, 3), "  (orange)"),
           hjust = 0, size = 3, color = "#cc5500") +
  labs(title    = "One sample: the OLS line versus the true population line",
       subtitle = paste0("The OLS estimate (", round(beta1_hat,3),
                         ") differs from the truth (", beta1_true,
                         ") due to the random noise u."),
       x = "x", y = "y") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

#### Step 4 --- Repeat 40 times: the spaghetti plot

```{r ols-step4, fig.height=3.8}
set.seed(42)
n_lines <- 40

# For each of 40 independent samples, run Steps 2-3 and save the coefficients.
# We only need the coefficients (b0, b1) -- not the raw data -- to draw the line.
all_coefs <- do.call(rbind, lapply(seq_len(n_lines), function(i) {
  x <- runif(n_obs, 0, 10)
  y <- beta0_true + beta1_true * x + rnorm(n_obs, 0, sigma_u)
  m <- lm(y ~ x)
  data.frame(sim = i, b0 = coef(m)[1], b1 = coef(m)["x"])
}))

cat("Summary of 40 OLS slope estimates:\n")
cat("  Mean:   ", round(mean(all_coefs$b1), 3),
    " (should be close to true beta1 =", beta1_true, ")\n")
cat("  Min:    ", round(min(all_coefs$b1), 3), "\n")
cat("  Max:    ", round(max(all_coefs$b1), 3), "\n")
cat("  Range:  ", round(diff(range(all_coefs$b1)), 3),
    " -- this is sampling variability\n")

# Build (x, y) prediction pairs for each of the 40 lines
x_grid <- seq(0, 10, length.out = 80)
df_lines <- do.call(rbind, lapply(seq_len(nrow(all_coefs)), function(i) {
  data.frame(sim = i, x = x_grid,
             y   = all_coefs$b0[i] + all_coefs$b1[i] * x_grid)
}))

ggplot() +
  geom_line(data = df_lines, aes(x = x, y = y, group = sim),
            color = "#2255aa", alpha = 0.2, linewidth = 0.45) +
  geom_abline(slope = beta1_true, intercept = beta0_true,
              color = "#cc2222", linewidth = 1.5) +
  annotate("text", x = 0.3, y = 32, color = "#cc2222", hjust = 0, size = 3,
           fontface = "bold", label = "True population line (red)") +
  annotate("text", x = 0.3, y = 29, color = "#2255aa", hjust = 0, size = 3,
           label = paste0(n_lines, " OLS estimates (blue)")) +
  coord_cartesian(ylim = c(-15, 40)) +
  labs(title    = "Each blue line is one OLS estimate from a different sample",
       subtitle = paste0("The cloud of lines centres on the true line -- this is unbiasedness: ",
                         "E[beta_hat_1] = beta_1"),
       x = "x", y = "y") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

#### Step 5 --- Repeat 2,000 times: the sampling distribution

```{r ols-step5, fig.height=3.8}
set.seed(1)
n_reps <- 2000

# Collect 2,000 slope estimates. replicate() runs the expression n_reps times
# and collects the result into a vector.
b1_all <- replicate(n_reps, {
  x <- runif(n_obs, 0, 10)
  y <- beta0_true + beta1_true * x + rnorm(n_obs, 0, sigma_u)
  coef(lm(y ~ x))["x"]   # return only the slope
})

cat("Results from", n_reps, "OLS estimates:\n")
cat("  Mean:                ", round(mean(b1_all), 4),
    " (should be", beta1_true, ")\n")
cat("  Standard deviation:  ", round(sd(b1_all), 4), "\n")

# Theoretical SE from Wooldridge Theorem 2.2:
# SE(beta_hat_1) = sigma / sqrt(SST_x)
# For x ~ Uniform(0,10): Var(x) = 10^2/12, E[SST_x] = n * Var(x)
se_theoretical <- sigma_u / sqrt(n_obs * 10^2 / 12)
cat("  Theoretical SE:      ", round(se_theoretical, 4),
    " (from Theorem 2.2: sigma/sqrt(E[SST_x]))\n")
cat("  Agreement:           ",
    round(abs(sd(b1_all) - se_theoretical), 4), "difference\n")
```

```{r ols-step5b, fig.height=3.8}
# Plot the empirical histogram and overlay the theoretical normal density
ggplot(data.frame(b1 = b1_all), aes(x = b1)) +
  geom_histogram(aes(y = after_stat(density)), bins = 55,
                 fill = "#2255aa", alpha = 0.7, color = "white") +
  # Theoretical density: N(beta1_true, se_theoretical^2)
  stat_function(fun  = dnorm,
                args = list(mean = beta1_true, sd = se_theoretical),
                color = "#cc2222", linewidth = 1.1) +
  geom_vline(xintercept = beta1_true, linetype = "dashed") +
  annotate("text", x = beta1_true + 0.07, y = 1.20,
           label = paste0("True beta_1 = ", beta1_true), hjust = 0, size = 3) +
  annotate("text", x = beta1_true + 0.07, y = 1.00,
           label = paste0("Theoretical SE = ", round(se_theoretical, 3)),
           hjust = 0, size = 3, color = "#cc2222") +
  annotate("text", x = beta1_true + 0.07, y = 0.80,
           label = paste0("Empirical SD   = ", round(sd(b1_all), 3)),
           hjust = 0, size = 3, color = "#2255aa") +
  labs(title    = paste0("Sampling distribution of beta_hat_1  (", n_reps, " replications)"),
       subtitle = paste0("Histogram = simulated;  red curve = N(beta_1, SE^2) from Theorem 2.2.  ",
                         "They match -- the theory is correct."),
       x = "OLS estimate of beta_1", y = "Density") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

---

### Confidence Intervals: What Does 95% Actually Mean?

**Goal:** construct 100 OLS confidence intervals and show concretely that approximately 95% of them capture the true $\beta_1 = 3$.

#### Step 1 --- How does `confint()` build the interval?

```{r ci-step1}
# Fit OLS to one sample and inspect what confint() computes
set.seed(55)
x_ex <- runif(n_obs, 0, 10)
y_ex <- beta0_true + beta1_true * x_ex + rnorm(n_obs, 0, sigma_u)
m_ex <- lm(y_ex ~ x_ex)

# Extract ingredients manually
beta1_hat_ex <- coef(m_ex)["x_ex"]
se_hat_ex    <- summary(m_ex)$coefficients["x_ex", "Std. Error"]
df_ex        <- m_ex$df.residual           # degrees of freedom = n - 2
t_crit_ex    <- qt(0.975, df = df_ex)      # t critical value (not z, because sigma unknown)

lower_manual <- beta1_hat_ex - t_crit_ex * se_hat_ex
upper_manual <- beta1_hat_ex + t_crit_ex * se_hat_ex

cat("Manual construction of the 95% CI:\n")
cat("  beta_hat_1 =", round(beta1_hat_ex, 4), "\n")
cat("  SE(beta_hat_1) =", round(se_hat_ex, 4), "\n")
cat("  t critical value (df =", df_ex, ") =", round(t_crit_ex, 4), "\n")
cat("  Margin of error = t * SE =", round(t_crit_ex * se_hat_ex, 4), "\n")
cat("  CI = [", round(lower_manual, 4), ",", round(upper_manual, 4), "]\n\n")

# Now compare with confint()
cf_auto <- confint(m_ex, "x_ex", level = 0.95)
cat("confint() result:\n")
cat("  CI = [", round(cf_auto[1], 4), ",", round(cf_auto[2], 4), "]\n")
cat("They match -- confint() is doing exactly beta_hat +/- t * SE.\n")
```

#### Step 2 --- Check if this specific CI captures the true value

```{r ci-step2}
# A CI [lower, upper] captures beta1_true if lower <= beta1_true <= upper.
captures_ex <- lower_manual <= beta1_true & beta1_true <= upper_manual

cat("True beta_1 =", beta1_true, "\n")
cat("CI = [", round(lower_manual,3), ",", round(upper_manual,3), "]\n")
cat("Does this CI capture the true value?", captures_ex, "\n")

if (captures_ex) {
  cat("-> YES. The interval contains the truth.\n")
} else {
  cat("-> NO. This is one of the ~5% that miss the truth.\n")
}
```

#### Step 3 --- Repeat for 100 independent samples

```{r ci-step3}
set.seed(123)
n_ci <- 100   # number of CIs to construct

# For each of 100 samples: draw data, fit OLS, extract CI, check coverage.
ci_results <- lapply(seq_len(n_ci), function(i) {
  x  <- runif(n_obs, 0, 10)
  y  <- beta0_true + beta1_true * x + rnorm(n_obs, 0, sigma_u)
  m  <- lm(y ~ x)
  cf <- confint(m, "x", level = 0.95)
  data.frame(
    sample   = i,
    est      = coef(m)["x"],   # point estimate
    lower    = cf[1],
    upper    = cf[2],
    captures = cf[1] <= beta1_true & beta1_true <= cf[2]
  )
})

df_ci <- do.call(rbind, ci_results)

# Summary
cat("Results across", n_ci, "samples:\n")
cat("  CIs that capture beta_1 = 3: ", sum(df_ci$captures), "\n")
cat("  CIs that miss  beta_1 = 3:  ", sum(!df_ci$captures), "\n")
cat("  Empirical coverage rate:      ", round(mean(df_ci$captures)*100, 1), "%\n")
cat("  Nominal  coverage rate:       95%\n")
```

#### Step 4 --- Visualise all 100 intervals

```{r ci-step4, fig.height=9.0}
# Sort by point estimate to make the plot readable
df_ci <- df_ci[order(df_ci$est), ]
df_ci$rank <- seq_len(nrow(df_ci))

ggplot(df_ci, aes(x = est, xmin = lower, xmax = upper, y = rank,
                  color = captures)) +
  geom_errorbarh(height = 0.3, linewidth = 0.65) +
  geom_point(size = 1.3) +
  geom_vline(xintercept = beta1_true, linewidth = 0.9) +
  scale_color_manual(
    values = c("TRUE"  = "#2255aa", "FALSE" = "#cc2222"),
    labels = c("TRUE"  = paste0("Captures true beta_1 = ", beta1_true),
               "FALSE" = paste0("Misses  true beta_1 = ", beta1_true)),
    name = NULL) +
  annotate("text", x = beta1_true + 0.12, y = 98,
           label = paste0("True beta_1 = ", beta1_true),
           hjust = 0, size = 3, fontface = "bold") +
  labs(
    title    = paste0("100 repeated 95% CIs for beta_hat_1  (true beta_1 = ", beta1_true, ")"),
    subtitle = paste0(sum(df_ci$captures), " intervals (blue) capture the truth  |  ",
                      sum(!df_ci$captures), " intervals (red) miss it  |  ",
                      "Empirical coverage = ", round(mean(df_ci$captures)*100,1), "%"),
    x = "95% CI for beta_1",
    y = "Sample (sorted by point estimate)"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"), legend.position = "top")
```

::: {.book-example}
<div class="header">Reading this plot</div>

Each horizontal bar is one confidence interval from one sample. The vertical black line is the true $\beta_1 = 3$. **Blue** bars cross the true line --- they capture the truth. **Red** bars miss it entirely.

Notice that the intervals vary in both **position** (because $\hat{\beta}_1$ varies across samples) and **width** (because $\text{SE}(\hat{\beta}_1) = \hat{\sigma}/\sqrt{\text{SST}_x}$ also varies). About 5 out of 100 miss --- consistent with 95% confidence.

The 95% is a property of the *method* (the factory), not of any single interval (the ring). Once a specific interval is computed, it either contains $\beta_1$ or it does not. There is no probability left for that specific interval.
:::

#### Step 5 --- Confirm coverage across different confidence levels

```{r ci-step5, fig.height=3.6}
set.seed(999)
conf_levels <- c(0.80, 0.90, 0.95, 0.99)

# For each confidence level, repeat 2,000 times and check coverage
coverage_table <- do.call(rbind, lapply(conf_levels, function(lev) {
  # replicate() runs the expression 2,000 times and returns a vector
  cvg <- replicate(2000, {
    x  <- runif(n_obs, 0, 10)
    y  <- beta0_true + beta1_true * x + rnorm(n_obs, 0, sigma_u)
    cf <- confint(lm(y ~ x), "x", level = lev)
    cf[1] <= beta1_true & beta1_true <= cf[2]
  })
  # Also record the average width of the CI at this level
  wid <- replicate(500, {
    x  <- runif(n_obs, 0, 10)
    y  <- beta0_true + beta1_true * x + rnorm(n_obs, 0, sigma_u)
    cf <- confint(lm(y ~ x), "x", level = lev)
    cf[2] - cf[1]
  })
  data.frame(level = lev, label = paste0(lev*100, "%"),
             coverage = mean(cvg), avg_width = mean(wid))
}))

# Print the table
cat("Nominal vs. Empirical Coverage:\n\n")
cat(sprintf("%-12s %-18s %-14s\n", "Level", "Empirical coverage", "Avg CI width"))
cat(strrep("-", 46), "\n")
for (i in seq_len(nrow(coverage_table))) {
  cat(sprintf("%-12s %-18s %-14s\n",
    coverage_table$label[i],
    paste0(round(coverage_table$coverage[i]*100,1), "%"),
    round(coverage_table$avg_width[i], 3)))
}

# Plot
ggplot(coverage_table, aes(x = label, y = coverage)) +
  geom_col(fill = "#2255aa", alpha = 0.8, width = 0.5) +
  geom_errorbar(aes(ymin = level, ymax = level), color = "#cc2222",
                linewidth = 1.3, width = 0.5) +
  geom_text(aes(label = paste0(round(coverage*100,1), "%\n(width=",
                               round(avg_width,2), ")")),
            vjust = -0.3, size = 3) +
  scale_y_continuous(limits = c(0, 1.13), labels = scales::percent) +
  labs(title    = "Higher confidence level = wider intervals = better coverage",
       subtitle = "Red tick = target coverage. Empirical (blue bars) tracks the target.",
       x = "Confidence level", y = "Empirical coverage (2,000 samples)") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

---

### Quick Reference: Every Function Used

```{r function-reference, eval=FALSE}
# PROBABILITY FUNCTIONS
pnorm(x, mean, sd)        # P(X <= x) for X ~ N(mean, sd^2)
qnorm(p, mean, sd)        # value x such that P(X <= x) = p
dnorm(x, mean, sd)        # density of N(mean, sd^2) at x
pt(x, df)                 # P(T <= x) for T ~ t_df
qt(p, df)                 # value x such that P(T <= x) = p

# RANDOM NUMBER GENERATION
set.seed(n)               # fix the random seed for reproducibility
rnorm(n, mean, sd)        # draw n values from N(mean, sd^2)
runif(n, min, max)        # draw n values from Uniform(min, max)
replicate(n, expr)        # run expr n times, collect results into a vector

# DATA MANIPULATION
do.call(rbind, list)      # stack a list of data frames row-by-row
lapply(X, FUN)            # apply FUN to each element of X, return a list
sapply(X, FUN)            # same but try to simplify to a vector

# OLS
lm(y ~ x)                 # fit OLS: minimise sum of squared residuals
coef(model)               # extract estimated coefficients (b0, b1, ...)
summary(model)            # full output: SE, t-values, p-values, R^2
confint(model, "x", 0.95) # 95% CI for the coefficient on x
                          # uses: beta_hat +/- t_{alpha/2, n-2} * SE(beta_hat)

# VECTORS OF LOGICALS
abs(Z) > z_crit           # TRUE where |Z| > 1.96 (rejection rule)
mean(logical_vector)      # proportion of TRUEs = empirical probability
sum(logical_vector)       # number of TRUEs
```

::: {.book-example}
<div class="header">Three ideas that connect everything in this tutorial</div>

**1. Randomness lives in the data, not in the parameters.**
The true $\beta_1 = 3$ is fixed. What changes across samples is $\hat{\beta}_1$ and $\text{SE}(\hat{\beta}_1)$ --- so the confidence interval (built from these) is what is random, not $\beta_1$.

**2. Size is a calibration guarantee, not an accuracy guarantee.**
$\alpha = 0.05$ means the testing procedure has a 5% false-alarm rate over many hypothetical experiments where $H_0$ is true. It says nothing about whether $H_0$ is true in your specific experiment.

**3. Power depends on the signal-to-noise ratio $\delta = |\beta^*|/\sigma_{\hat{\beta}}$.**
$\delta$ grows when the true effect is large *or* when we have more/better data (smaller SE). Both routes lead to higher power. This is why large datasets are needed to detect small economic effects.
:::

---

## Part 2: Problem Set --- Concepts and Exercises (50 min)

---

### Block 1: Type I and Type II Errors (10 min)

::: {.book-example}
<div class="header">The Four Possible Outcomes of a Hypothesis Test</div>

Every hypothesis test makes one binary decision --- **Reject $H_0$** or **Fail to reject $H_0$** --- while the truth is in one of two states. This gives four possible outcomes:

| | **$H_0$ is actually TRUE** | **$H_0$ is actually FALSE** |
|:---|:---:|:---:|
| **Reject $H_0$** | **Type I Error** (false positive) | **Correct!** (Power) |
| **Fail to reject $H_0$** | **Correct!** (true negative) | **Type II Error** (false negative) |

**Type I Error ($\alpha$):** We reject $H_0$ when it is true. We see a pattern that is not there.

**Type II Error ($\beta$):** We fail to reject $H_0$ when it is false. A real effect is present but we miss it.

**The asymmetry:** In most contexts the two errors are not equally costly. Criminal law ("beyond reasonable doubt") prioritises minimising Type I --- convicting an innocent person --- even at the cost of more Type II errors. Medical screening may reverse this priority.
:::

**Quick Exercise.** Fill in the table for the following context:

*We test whether a job-training programme has an effect on wages. $H_0$: the programme has **no effect** on wages.*

| | **Type I Error** | **Type II Error** |
|:---|:---|:---|
| **What happens** | | |
| **In words** | | |
| **Real-world consequence** | | |

<details class="solution">
<summary>Solution</summary>

**$H_0$:** The programme has no effect ($\beta_1 = 0$).

**Type I Error:** We reject $H_0$ and conclude the programme works --- but it actually does not. We credit a useless programme and waste public resources.

**Type II Error:** We fail to reject $H_0$ and conclude the programme is ineffective --- but it actually works. A valuable policy is dismissed and workers miss the benefit.

**Which is worse?** Depends on the stakes. If the programme is cheap to run, Type II is worse (we lose real gains). If funding is scarce, Type I is worse (we fund an ineffective policy).
</details>

---

::: {.book-example}
<div class="header">How to read the figure below</div>

The figure has **two panels**, one for each state of the world.

**Left panel --- $H_0$ is true (programme has no effect, $\beta_1 = 0$).**
The bell curve shows all values of the test statistic $Z = \hat{\beta}_1 / \sigma_{\hat{\beta}_1}$ we could observe across many repeated studies, when there truly is no effect. The curve is centred at zero. The dashed vertical lines at $\pm 1.96$ are the critical values: we reject $H_0$ whenever $Z$ falls outside them.

- **Red tails (Type I Error, 5%):** these are the studies where we *would* reject $H_0$ --- even though it is true. The programme does nothing, but by bad luck our estimate lands far from zero. This happens 5% of the time. We cannot eliminate it; we chose $\alpha = 0.05$ precisely to accept it.
- **White centre (95%):** the studies where we correctly do not reject $H_0$.

**Right panel --- $H_0$ is false (programme raises wages by $\beta_1^* = 3$).**
Now the true effect is non-zero, so the distribution of $Z$ shifts to the right --- it is no longer centred at zero but at $\delta = \beta_1^* / \sigma_{\hat{\beta}_1} = 1.5$.

- **Green area (Power):** studies where $Z$ exceeds the critical value and we *correctly* reject $H_0$. The programme works, and we detect it.
- **Grey area (Type II Error):** studies where $Z$ stays inside the critical values and we *fail* to detect the real effect. The distribution has shifted, but not enough for most draws to clear the bar.

**Key intuition:** the two panels share the same critical values (dashed lines). What changes is the *centre* of the bell curve. When the true effect is small, the right-panel curve barely moves --- most of its mass stays inside the critical values, giving low power and high Type II error. A larger true effect shifts the curve further right, putting more mass beyond the threshold.
:::

```{r error-types-diagram, echo=FALSE, fig.height=4.0}
# Dense grid of Z values
z_seq  <- seq(-5, 8, length.out = 1000)
z_crit <- qnorm(0.975)   # 1.96: the value splitting the central 95% from the tails
delta  <- 1.5            # delta = beta* / sigma_hat = 3 / 2

# Sampling distribution of Z under H0 (centred at 0) and under H1 (shifted by delta)
df_h0  <- data.frame(z = z_seq, density = dnorm(z_seq, 0, 1),
                     scenario = "H0 true  (beta = 0)")
df_h1  <- data.frame(z = z_seq, density = dnorm(z_seq, delta, 1),
                     scenario = "H0 false  (beta* = 3, delta = 1.5)")
df_all <- rbind(df_h0, df_h1)

# Text labels placed inside each coloured region
labels_df <- data.frame(
  z        = c(2.7,   0.0,   3.3,   0.5),
  density  = c(0.06,  0.10,  0.11,  0.09),
  label    = c("Type I\nError\n(5%)", "Correctly\nnot rejected\n(95%)",
               "Power\n(32%)", "Type II\nError\n(68%)"),
  color_lbl = c("#cc2222", "grey30", "#2a7a2a", "grey50"),
  scenario = c("H0 true  (beta = 0)", "H0 true  (beta = 0)",
               "H0 false  (beta* = 3, delta = 1.5)",
               "H0 false  (beta* = 3, delta = 1.5)")
)

ggplot(df_all, aes(x = z, y = density)) +
  geom_line(linewidth = 0.9, color = "grey30") +
  geom_ribbon(data = df_h0 %>% filter(z >  z_crit),
              aes(ymin = 0, ymax = density), fill = "#cc2222", alpha = 0.5) +
  geom_ribbon(data = df_h0 %>% filter(z < -z_crit),
              aes(ymin = 0, ymax = density), fill = "#cc2222", alpha = 0.5) +
  geom_ribbon(data = df_h1 %>% filter(z > z_crit),
              aes(ymin = 0, ymax = density), fill = "#2a7a2a", alpha = 0.55) +
  geom_ribbon(data = df_h1 %>% filter(z <= z_crit & z >= -z_crit),
              aes(ymin = 0, ymax = density), fill = "grey60", alpha = 0.5) +
  geom_vline(xintercept = c(-z_crit, z_crit), linetype = "dashed") +
  geom_text(data = labels_df,
            aes(x = z, y = density, label = label),
            color = labels_df$color_lbl, size = 2.6, inherit.aes = FALSE) +
  facet_wrap(~scenario) +
  labs(title    = "Type I Error, Type II Error, and Power",
       subtitle = paste0("Two-sided test, alpha = 0.05  |  critical values: +/-",
                         round(z_crit, 2), "  |  beta* = 3, sigma_hat = 2, delta = 1.5"),
       x = "Test statistic  Z = beta_hat / sigma_hat", y = "Density") +
  theme_tutorial()
```

---

### Block 2: Size (7 min)

::: {.book-example}
<div class="header">Definition of Size</div>

$$\text{Size} = P(\text{reject } H_0 \mid H_0 \text{ is true}) = \alpha$$

The **size** of a test equals the Type I error probability. When we set $\alpha = 0.05$ and use critical value $z_{0.025} = 1.96$, we construct a procedure that --- **when $H_0$ is true** --- rejects exactly 5% of the time in repeated sampling.

$$P(|Z| > 1.96 \mid H_0) = P(Z > 1.96) + P(Z < -1.96) = 0.025 + 0.025 = 0.05$$

**What $\alpha = 0.05$ means:** if we ran the same experiment 1,000 times in a world where $H_0$ is true, we would falsely reject $H_0$ about 50 times --- purely by random chance.

**What $\alpha$ does NOT mean:** it is not the probability that $H_0$ is true. It is a long-run frequency property of the *procedure*, not of any single conclusion.
:::

---

### Block 3: Power (8 min)

::: {.book-example}
<div class="header">Definition of Power</div>

$$\text{Power}(\beta^*) = P(\text{reject } H_0 \mid \beta = \beta^*) = 1 - P(\text{Type II Error} \mid \beta = \beta^*)$$

Power is the probability of **correctly** rejecting $H_0$ when the true parameter is $\beta^*$.

**What makes power high?**

1. Large true effect $|\beta^*|$: the bigger the signal, the easier to detect.
2. Small $\sigma_{\hat{\beta}}$: more precision means larger signal-to-noise ratio.
3. Larger $\alpha$: a looser Type I threshold makes rejection easier --- at a cost.

**Analytical formula.** When $Z = \hat{\beta}/\sigma_{\hat{\beta}}$ and the true value is $\beta^*$:

$$\text{Power}(\beta^*) = \Phi\!\left(\frac{|\beta^*|}{\sigma_{\hat{\beta}}} - z_{\alpha/2}\right) + \Phi\!\left(-\frac{|\beta^*|}{\sigma_{\hat{\beta}}} - z_{\alpha/2}\right) \approx \Phi\!\left(\delta - 1.96\right) \quad\text{for large }\delta$$

where $\delta = |\beta^*|/\sigma_{\hat{\beta}}$ is the **signal-to-noise ratio**.

**Key property:** Power is a *function* of $\beta^*$, not a single number. When $\beta^* = 0$ (i.e., $H_0$ is exactly true), Power $= \alpha$. As $|\beta^*|$ grows, Power increases toward 1.
:::

::: {.book-example}
<div class="header">How to read the power curve below</div>

The x-axis is the **true effect** $\beta^*$ --- a value we never observe in practice. The y-axis is the **probability that our test detects it**.

Read the curve like this: "*If the true wage effect of the programme were $\beta^*$ dollars/hour, how often would our test correctly reject $H_0$?*"

- **At $\beta^* = 0$:** $H_0$ is exactly true. Power $= \alpha = 5\%$. We reject only by chance --- this is the Type I error floor. The curve never goes below this line.
- **At $\beta^* = 3$:** The programme raises wages by \$3/hour. Power $\approx 32\%$. We catch the effect only 1 in 3 times --- low power because 3 is only 1.5 standard errors from zero.
- **At $\beta^* = 6$:** Power $\approx 85\%$. The effect is now 3 standard errors from zero --- large enough to detect reliably.
- **The curve is symmetric:** a negative effect of $-3$ is just as hard (or easy) to detect as $+3$.
- **The 80% dashed line** is a conventional benchmark for "adequate power" used in study design.

**Why does the curve rise as $|\beta^*|$ grows?** Because a larger true effect shifts the distribution of $Z$ further from zero (larger $\delta = |\beta^*|/\sigma_{\hat\beta}$), placing more probability mass beyond the critical values $\pm 1.96$. In the error-types diagram above, this is equivalent to moving the right-panel bell curve further to the right.
:::

```{r power-function, echo=FALSE, fig.height=3.8}
sigma_hat <- 2
z_crit    <- qnorm(0.975)
beta_grid <- seq(-8, 8, by = 0.1)
delta     <- beta_grid / sigma_hat

# Analytical power at each point: P(|Z| > 1.96) when Z ~ N(delta, 1)
power_analytic <- pnorm(-z_crit - delta) + (1 - pnorm(z_crit - delta))
df_pw <- data.frame(beta = beta_grid, power = power_analytic)

ggplot(df_pw, aes(x = beta, y = power)) +
  geom_line(color = "#2255aa", linewidth = 1.1) +
  geom_hline(yintercept = 0.05, linetype = "dotted", color = "grey50") +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "#2a7a2a", linewidth = 0.8) +
  geom_vline(xintercept = 0, color = "grey40", linewidth = 0.4) +
  annotate("point", x = c(3, 6), y = c(0.323, 0.851), color = "#cc2222", size = 3) +
  annotate("text",  x = c(3.3, 6.3), y = c(0.323, 0.851),
           label = c("beta*=3  =>  Power = 32%", "beta*=6  =>  Power = 85%"),
           hjust = 0, size = 2.8, color = "#cc2222") +
  annotate("text", x = 7.8, y = 0.08, label = "floor = alpha = 5%",
           hjust = 1, size = 2.8, color = "grey50") +
  annotate("text", x = 7.8, y = 0.83, label = "80% benchmark",
           hjust = 1, size = 2.8, color = "#2a7a2a") +
  annotate("text", x = 0.15, y = 0.12, label = "beta*=0:\nPower=alpha",
           hjust = 0, size = 2.5, color = "grey40") +
  scale_y_continuous(labels = scales::percent, breaks = seq(0, 1, 0.2)) +
  labs(title    = "Power function: for each possible true effect, how often do we detect it?",
       subtitle = "sigma_hat = 2, alpha = 0.05.  Red dots = the two values computed in Exercise 1.",
       x = "True wage effect  beta*  ($/hour)", y = "Power  =  P(reject H0)") +
  theme_tutorial()
```

---

### Exercise: A Complete Example (25 min)

::: {.book-example}
<div class="header">The Setting</div>

A researcher wants to know whether a job-training programme raises hourly wages. She estimates the effect with OLS and obtains $\hat{\beta}_1$ (dollars/hour). Under SLR.1--SLR.6 the estimator is normally distributed:

$$\hat{\beta}_1 \sim N\!\bigl(\beta_1,\; \sigma_{\hat{\beta}_1}^2\bigr) \qquad \text{with known } \sigma_{\hat{\beta}_1} = 2 \text{ \$/hour.}$$

She tests $H_0\colon \beta_1 = 0$ (no effect) against $H_1\colon \beta_1 \neq 0$ at $\alpha = 0.05$.

The test statistic is $Z = \hat{\beta}_1\,/\,2$, which follows $N(0,1)$ under $H_0$.
:::

#### Part A --- Theoretical Calculations

**(a) Decision rule and critical values.** Derive the values of $\hat{\beta}_1$ that lead to rejection of $H_0$.

<details class="solution">
<summary>Solution</summary>

Under $H_0$, the test statistic $Z = \hat{\beta}_1 / \sigma_{\hat{\beta}_1}$ follows a standard normal distribution:

$$Z = \frac{\hat{\beta}_1}{2} \sim N(0,1) \quad \text{when } \beta_1 = 0.$$

We reject $H_0$ at $\alpha = 0.05$ whenever $Z$ falls in the rejection region:

$$|Z| > z_{\alpha/2} = z_{0.025} = 1.96$$

Translating back to $\hat{\beta}_1$:

$$\left|\frac{\hat{\beta}_1}{2}\right| > 1.96 \;\iff\; |\hat{\beta}_1| > 1.96 \times 2 = 3.92 \text{ \$/hour}$$

**Decision rule:** reject $H_0$ if $\hat{\beta}_1 > 3.92$ or $\hat{\beta}_1 < -3.92$.
</details>

**(b) Size and Type I error.** What is the size of this test? Compute it explicitly and interpret in context.

<details class="solution">
<summary>Solution</summary>

The size is the probability of Type I error: rejecting $H_0$ when it is true ($\beta_1 = 0$). Since $Z \sim N(0,1)$ under $H_0$:

$$\text{Size} = P(|Z| > 1.96 \mid \beta_1 = 0) = P(Z > 1.96) + P(Z < -1.96) = 0.025 + 0.025 = \mathbf{0.05}$$

*In context:* if the training programme truly has no effect on wages, and we ran this study 100 times with different random samples, we would wrongly conclude that it works in approximately 5 of those studies --- purely due to random variation in the data. We tolerate this 5% false-positive rate by choice ($\alpha = 0.05$).
</details>

**(c) Power and Type II error.** Suppose the programme *actually* raises wages by $\beta_1^* = 3$ \$/hour.

1. Compute the **power** step by step.
2. Compute the **Type II error probability**.
3. Interpret both numbers in context.

<details class="solution">
<summary>Solution</summary>

**(i) Power --- step by step.**

*Step 1.* When the truth is $\beta_1^* = 3$, the OLS estimator has distribution:

$$\hat{\beta}_1 \sim N(3,\; 4) \quad \Rightarrow \quad Z = \frac{\hat{\beta}_1}{2} \sim N\!\left(\frac{3}{2},\; 1\right) = N(1.5,\;1)$$

The test statistic is no longer centred at zero --- it is shifted right by $\delta = \beta_1^*/\sigma_{\hat{\beta}_1} = 3/2 = 1.5$.

*Step 2.* Power is the probability of falling in the rejection region under this shifted distribution:

$$\text{Power}(\beta_1^* = 3) = P(Z > 1.96 \mid Z \sim N(1.5,1)) + P(Z < -1.96 \mid Z \sim N(1.5,1))$$

*Step 3.* Standardise each term. If $Z \sim N(1.5, 1)$, then $W = Z - 1.5 \sim N(0,1)$, so:

$$P(Z > 1.96) = P(W > 1.96 - 1.5) = P(W > 0.46) = 1 - \Phi(0.46) = 1 - 0.677 = 0.323$$

$$P(Z < -1.96) = P(W < -1.96 - 1.5) = P(W < -3.46) = \Phi(-3.46) \approx 0.0003$$

*Step 4.* Combine:

$$\text{Power}(3) = 0.323 + 0.0003 \approx \mathbf{0.323 = 32.3\%}$$

**(ii) Type II error.** The Type II error is the complement of power (in the column "$H_0$ false"):

$$P(\text{Type II}) = 1 - \text{Power}(3) = 1 - 0.323 = \mathbf{0.677 = 67.7\%}$$

**(iii) Interpretation.** The programme raises wages by \$3/hour, but our test detects this real effect only 32% of the time. In 68% of studies we would fail to reject $H_0$ and wrongly conclude the programme is ineffective --- a Type II error. The power is low because \$3/hour is only 1.5 standard errors away from zero; the two hypotheses are hard to tell apart with this level of precision.
</details>

**(d) Summary table.** Fill in all four cells with the **probabilities** for this specific example ($\sigma_{\hat{\beta}_1} = 2$, $\alpha = 0.05$, $\beta_1^* = 3$).

| | **$H_0$ true** ($\beta_1 = 0$) | **$H_0$ false** ($\beta_1^* = 3$) |
|:---|:---:|:---:|
| **Reject $H_0$** | | |
| **Fail to reject $H_0$** | | |

<details class="solution">
<summary>Solution</summary>

| | **$H_0$ true** ($\beta_1 = 0$) | **$H_0$ false** ($\beta_1^* = 3$) |
|:---|:---:|:---:|
| **Reject $H_0$** | **Type I error** $= \alpha = 0.05$ | **Power** $= 0.323$ |
| **Fail to reject $H_0$** | Correct $= 1 - \alpha = 0.95$ | **Type II error** $= 1 - 0.323 = 0.677$ |

**Key relationship:** each column sums to 1 (the two outcomes are exhaustive). Type I error and Power live in *different columns* --- they describe different states of the world. Reducing $\alpha$ (tighter Type I control) pushes the critical values outward, which makes it harder to reject $H_0$ and therefore *increases* Type II error. The only way to reduce both errors simultaneously is to increase the precision of $\hat{\beta}_1$ (i.e., larger sample size $n$).
</details>

**(e) What if the true effect were larger?** Compute power when $\beta_1^* = 6$ \$/hour, showing all steps. Explain why power is higher than in (c).

<details class="solution">
<summary>Solution</summary>

*Step 1.* New signal-to-noise ratio: $\delta = 6/2 = 3$. The distribution of $Z$ is now $N(3, 1)$.

*Step 2--3.* Standardise:

$$P(Z > 1.96 \mid Z \sim N(3,1)) = P(W > 1.96 - 3) = P(W > -1.04) = \Phi(1.04) = 0.851$$

$$P(Z < -1.96 \mid Z \sim N(3,1)) = P(W < -1.96 - 3) = \Phi(-4.96) \approx 0$$

*Step 4.*

$$\text{Power}(6) = 0.851 + 0 \approx \mathbf{0.851 = 85.1\%}$$

Type II error $= 1 - 0.851 = 14.9\%$.

*Why is power higher?* When $\beta_1^* = 6$, the distribution of $Z$ is centred at $\delta = 3$ --- three standard errors from zero. Most of its mass now lies to the right of the critical value $1.96$, making rejection the most likely outcome. Compared to $\beta_1^* = 3$ (where $\delta = 1.5$), the distribution has moved further from the threshold, and detection is much easier.
</details>

---

#### Part B --- Simulation

::: {.book-example}
<div class="header">Goal</div>

We verify the numbers from Part A by simulation using the same job-training context.

**The idea:** imagine running the wage study 10,000 times --- drawing a new random sample of workers each time and computing $\hat{\beta}_1$. We do this twice: once in a world where the programme has no effect (**Scenario A**: $\beta_1 = 0$) and once where it truly raises wages by \$3/hour (**Scenario B**: $\beta_1^* = 3$).

In each simulated study we apply the same decision rule: reject $H_0$ if $|\hat{\beta}_1| > 3.92$.

- In Scenario A, the fraction of rejections estimates the **size** (should be $\approx 0.05$).
- In Scenario B, the fraction of rejections estimates the **power** (should be $\approx 0.323$).
- In Scenario B, the fraction of *non*-rejections estimates the **Type II error** (should be $\approx 0.677$).
:::

```{r simulation}
set.seed(42)
sigma_hat <- 2            # SE of beta_hat (same as Part A)
z_crit    <- qnorm(0.975) # 1.96  =>  reject when |beta_hat| > 1.96 * 2 = 3.92
n_sim     <- 10000        # number of simulated studies

# ---------------------------------------------------------------
# Scenario A: H0 is TRUE -- programme has NO effect (beta_1 = 0)
# Each draw is one hypothetical study: a new sample of workers,
# a new OLS estimate.  Here beta_hat ~ N(0, sigma_hat^2).
# ---------------------------------------------------------------
bhat_A   <- rnorm(n_sim, mean = 0, sd = sigma_hat)
reject_A <- abs(bhat_A / sigma_hat) > z_crit   # TRUE = wrongly rejected H0

cat("--- Scenario A: programme has NO effect (beta_1 = 0) ---\n")
cat("  Reject H0 (Type I error):  ", round(mean(reject_A),  3),
    "  [theory: 0.050]\n")
cat("  Do not reject (correct):   ", round(mean(!reject_A), 3),
    "  [theory: 0.950]\n\n")

# ---------------------------------------------------------------
# Scenario B: H0 is FALSE -- true effect is beta_1* = 3 $/hour
# Now beta_hat ~ N(3, sigma_hat^2).  The distribution has shifted.
# ---------------------------------------------------------------
bhat_B   <- rnorm(n_sim, mean = 3, sd = sigma_hat)
reject_B <- abs(bhat_B / sigma_hat) > z_crit   # TRUE = correctly rejected H0

cat("--- Scenario B: programme raises wages by $3/hour (beta_1* = 3) ---\n")
cat("  Reject H0 (Power):         ", round(mean(reject_B),  3),
    "  [theory: 0.323]\n")
cat("  Do not reject (Type II):   ", round(mean(!reject_B), 3),
    "  [theory: 0.677]\n")
```

```{r sim-plot, echo=FALSE, fig.height=4.2}
z_seq   <- seq(-5, 7, length.out = 1000)
z_crit2 <- qnorm(0.975)

df_A <- data.frame(z = z_seq, dens = dnorm(z_seq, 0,   1),
                   scenario = "Scenario A: H0 true (beta=0)\nNo wage effect")
df_B <- data.frame(z = z_seq, dens = dnorm(z_seq, 3/2, 1),
                   scenario = "Scenario B: H0 false (beta*=3)\nProgramme raises wages $3/hr")
df_both <- rbind(df_A, df_B)

ldf <- data.frame(
  z    = c( 2.6,  0.0,  3.5,  0.5),
  dens = c(0.06, 0.09, 0.11, 0.085),
  lbl  = c("Type I\nerror\n5%", "Correct\n95%", "Power\n32%", "Type II\nerror\n68%"),
  col  = c("#cc2222", "grey30", "#2a7a2a", "grey50"),
  scenario = c(
    "Scenario A: H0 true (beta=0)\nNo wage effect",
    "Scenario A: H0 true (beta=0)\nNo wage effect",
    "Scenario B: H0 false (beta*=3)\nProgramme raises wages $3/hr",
    "Scenario B: H0 false (beta*=3)\nProgramme raises wages $3/hr"
  )
)

ggplot(df_both, aes(x = z, y = dens)) +
  geom_line(linewidth = 0.9, color = "grey30") +
  geom_ribbon(data = subset(df_A, z >  z_crit2),
              aes(ymin=0, ymax=dens), fill="#cc2222", alpha=0.5) +
  geom_ribbon(data = subset(df_A, z < -z_crit2),
              aes(ymin=0, ymax=dens), fill="#cc2222", alpha=0.5) +
  geom_ribbon(data = subset(df_B, z >  z_crit2),
              aes(ymin=0, ymax=dens), fill="#2a7a2a", alpha=0.55) +
  geom_ribbon(data = subset(df_B, z >= -z_crit2 & z <= z_crit2),
              aes(ymin=0, ymax=dens), fill="grey60", alpha=0.5) +
  geom_vline(xintercept = c(-z_crit2, z_crit2), linetype="dashed") +
  geom_text(data = ldf, aes(x=z, y=dens, label=lbl),
            color=ldf$col, size=2.6, inherit.aes=FALSE) +
  facet_wrap(~scenario) +
  labs(title    = "Simulation confirms the theory: same numbers, same decision rule",
       subtitle = "Left: size = 5% (red = false positives).  Right: power = 32% (green) and Type II = 68% (grey).",
       x = "Test statistic  Z = beta_hat / 2", y = "Density") +
  theme_tutorial()
```

<details class="solution">
<summary>Solution</summary>

**Reading the results:**

- **Scenario A --- size $\approx 0.05$:** in 10,000 studies where the programme has no effect, we wrongly rejected $H_0$ in about 5% of them. This is exactly the Type I error rate we chose ($\alpha = 0.05$). The result confirms that our critical value of $\pm 1.96$ is correctly calibrated.

- **Scenario B --- power $\approx 0.323$:** in 10,000 studies where the programme genuinely raises wages by \$3/hour, we correctly detected the effect in only about 32% of them --- matching the analytical calculation in Part A(c). This is low power: a real, meaningful wage gain goes undetected most of the time.

- **Scenario B --- Type II error $\approx 0.677$:** in the remaining 68% of Scenario B studies, we failed to reject $H_0$ even though the programme works. These are false negatives --- studies that would lead a policy-maker to cancel a beneficial programme.

- **Connection to the figure:** the two panels of the plot correspond exactly to the two scenarios. The left panel (Scenario A) shows the $N(0,1)$ distribution of $Z$; the red tails are the 5% false-positive rate. The right panel (Scenario B) shows the shifted $N(1.5,1)$ distribution; the green area is power and the grey area is the Type II error.
</details>
