---
title: "Tutorial 8 --- Exercise 2: The $F$-Test in R"
author: "Introductory Econometrics --- TA Session"
date: "Duration: 50 minutes | Based on Exercise 1"
output:
  pdf_document:
    latex_engine: pdflatex
    keep_tex: false
    number_sections: false
geometry: margin=2.3cm
header-includes:
  - \input{preamble_t8.tex}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead[L]{\small Tutorial 8 --- Exercise 2}
  - \fancyhead[R]{\small Introductory Econometrics}
  - \setlength{\parskip}{5pt}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 6.3, fig.height = 3.8,
  out.width = "\\textwidth"
)
```

In Exercise 1 you derived the $F$-statistic on paper. Now you will compute it from scratch in R, verify it against built-in functions, and see the test make the right decision.

We simulate data from the same wage equation as Lecture 14. The key advantage of simulation: because we choose the true parameters, we *know* whether $H_0$ is true or false --- something we never know with real data. This lets us check whether the test behaves as the theory predicts.

---

**Step 1. Generate the data.**

We create a dataset of $n = 500$ workers with the same variables as the Lecture 14 wage equation.

```{r dgp}
rm(list = ls())       # start with a clean environment
set.seed(326)         # fix the random seed so results are reproducible
n <- 500              # sample size: 500 workers
```

First, we generate the regressors. Each line creates one variable for all $n$ individuals at once:

```{r regressors}
# runif(n, a, b) draws n values uniformly between a and b
Exp      <- runif(n, 1, 30)       # current job experience: between 1 and 30 years
Exp2     <- Exp^2                  # experience squared (for diminishing returns)

PrevExp  <- runif(n, 0, 20)       # previous experience: between 0 and 20 years
PrevExp2 <- PrevExp^2              # previous experience squared

# sample(x, n, replace = TRUE) draws n integers from the set x with replacement
Educ     <- sample(8:20, n, replace = TRUE)   # education: 8 to 20 years
```

Next, we set the **true population parameters**. This is the "God's-eye view" --- we decide exactly how wages are determined. In real research you never know these; here we choose them so we can verify the test.

```{r parameters}
b0 <- 0.5             # intercept
b1 <- 0.04            # return to current experience (positive)
b2 <- -0.0008         # diminishing returns to experience (negative quadratic)
b3 <- 0               # <-- previous experience linear effect: ZERO
b4 <- 0               # <-- previous experience quadratic effect: ZERO
b5 <- 0.09            # return to education
sigma <- 0.4          # standard deviation of the error term
```

\begin{simbox}{Design choice: $H_0$ is TRUE by construction}
We set $\beta_3 = 0$ and $\beta_4 = 0$. This means previous experience has \textbf{no effect whatsoever} on wages in the population. The null hypothesis $H_0\colon \beta_3 = 0,\; \beta_4 = 0$ is true.

\medskip
\textbf{What should we expect?} If the $F$-test works correctly, it should \emph{fail to reject} $H_0$ about 95\% of the time (at $\alpha = 0.05$). The other 5\% of the time, it will reject even though $H_0$ is true --- that is a Type I error, and it happens by design. The whole point of setting $\alpha = 0.05$ is accepting a 5\% false rejection rate.
\end{simbox}

Finally, generate the outcome variable:

```{r outcome}
# rnorm(n, mean, sd) draws n values from a normal distribution N(mean, sd^2)
# Each draw represents the unobserved factors affecting worker i's wage
U <- rnorm(n, mean = 0, sd = sigma)

# The data-generating process: this is the TRUE model
# Since b3 = b4 = 0, the PrevExp terms contribute nothing
lnWage <- b0 + b1*Exp + b2*Exp2 + b3*PrevExp + b4*PrevExp2 + b5*Educ + U

# Combine everything into a data frame for use with lm()
wages <- data.frame(lnWage, Exp, Exp2, PrevExp, PrevExp2, Educ)
```

---

**Step 2. Estimate the unrestricted model and extract $SSR_{ur}$.**

The unrestricted model includes all five regressors --- it does not impose $H_0$. This is the full model from Exercise 1, Step 2.

```{r unrestricted}
unrestricted <- lm(lnWage ~ Exp + Exp2 + PrevExp + PrevExp2 + Educ,
                   data = wages)
summary(unrestricted)
```

\begin{bookexample}
\textbf{\color{blockcolor}Reading the output}

\medskip
Look at the coefficients on \texttt{PrevExp} and \texttt{PrevExp2}. Since $\beta_3 = \beta_4 = 0$ in the DGP, their estimated coefficients should be close to zero (but not exactly zero --- there is always sampling noise). Their individual $p$-values may or may not be below 0.05. Remember from Exercise 1, part (a): individual $t$-tests are \emph{not} the right way to test a joint hypothesis.
\end{bookexample}

Now extract $SSR_{ur}$. The function `resid()` returns the vector of OLS residuals $\hat{U}_i = Y_i - \hat{Y}_i$ for each observation. Squaring each one and summing gives $SSR_{ur} = \sum_{i=1}^{n} \hat{U}_i^2$:

```{r ssr-ur}
SSR_ur <- sum(resid(unrestricted)^2)     # sum of squared residuals
df_ur  <- unrestricted$df.residual       # residual df = n - k - 1 = 500 - 5 - 1 = 494

cat("SSR_ur =", round(SSR_ur, 4), "\n")
cat("Residual df (n - k - 1) =", df_ur, "\n")
```

---

**Step 3. Estimate the restricted model and extract $SSR_r$.**

Under $H_0\colon \beta_3 = 0,\; \beta_4 = 0$, we drop `PrevExp` and `PrevExp2` from the model. This is the restricted model from Exercise 1, Step 2.

```{r restricted}
restricted <- lm(lnWage ~ Exp + Exp2 + Educ, data = wages)

SSR_r <- sum(resid(restricted)^2)
cat("SSR_r =", round(SSR_r, 4), "\n")
```

Now verify the key property from Exercise 1, part (b): imposing restrictions can never improve the fit.

```{r sanity}
cat("SSR_r >= SSR_ur?", SSR_r >= SSR_ur, "\n")
cat("Fit cost (SSR_r - SSR_ur) =", round(SSR_r - SSR_ur, 4), "\n")
```

\begin{bookexample}
\textbf{\color{blockcolor}Interpreting the fit cost}

\medskip
The number $SSR_r - SSR_{ur}$ is the \textbf{total increase in the sum of squared residuals} when we drop $PrevExp$ and $PrevExp^2$. Since $H_0$ is true in our DGP (these variables genuinely have zero coefficients), the fit cost should be \textbf{small} --- just sampling noise. If this number were large, it would suggest the excluded variables were genuinely important.
\end{bookexample}

---

**Step 4. Compute the $F$-statistic by hand.**

Now we apply the formula from Exercise 1, Step 3:
$$F = \frac{(SSR_r - SSR_{ur})\,/\,q}{SSR_{ur}\,/\,(n - k - 1)}$$

```{r f-stat}
q <- 2    # number of restrictions (beta_3 = 0 AND beta_4 = 0)

# Numerator: average fit cost per restriction
numerator <- (SSR_r - SSR_ur) / q

# Denominator: mean squared error from unrestricted model (estimate of sigma^2)
denominator <- SSR_ur / df_ur

# F-statistic: signal-to-noise ratio
F_manual <- numerator / denominator

cat("Numerator (fit cost per restriction):", round(numerator, 4), "\n")
cat("Denominator (MSE, estimate of sigma^2):", round(denominator, 4), "\n")
cat("F-statistic:", round(F_manual, 4), "\n")
```

\begin{bookexample}
\textbf{\color{blockcolor}What does this number mean?}

\medskip
The $F$-statistic tells you how many times larger the fit loss per restriction is compared to the background noise. If $F \approx 1$, the fit loss is about what you would expect from random variation alone --- no evidence against $H_0$. If $F$ is much larger than 1, the fit loss is too big to be explained by chance. In our DGP where $H_0$ is true, $F$ should be modest (close to 1 on average).
\end{bookexample}

---

**Step 5. Compute the $p$-value and make a decision.**

We need two R functions for the $F$-distribution. They work exactly like `pnorm`/`qnorm` from Tutorial 7, but for the $F$-distribution instead of the normal:

\begin{bookexample}
\textbf{\color{blockcolor}Two essential R functions}

\medskip
\textbf{\texttt{qf(p, df1, df2)}} answers: ``what $F$-value has fraction $p$ of the $F_{df1, df2}$ distribution below it?'' This gives the \textbf{critical value}.

\medskip
Example: \texttt{qf(0.95, df1 = 2, df2 = 494)} returns the value such that 95\% of the $F_{2,494}$ distribution lies below it. If your $F$-statistic exceeds this value, you reject $H_0$ at $\alpha = 0.05$.

\medskip
\textbf{\texttt{pf(x, df1, df2, lower.tail = FALSE)}} answers: ``what fraction of the $F_{df1, df2}$ distribution lies \emph{above} $x$?'' This gives the \textbf{$p$-value}.

\medskip
\warn{Always use \texttt{lower.tail = FALSE}.} Without it, \texttt{pf()} returns the area \emph{below} $x$, which is the wrong tail. A large $F$ would give a $p$-value near 1 instead of near 0.
\end{bookexample}

```{r decision}
alpha  <- 0.05

# Critical value: the 95th percentile of the F(2, 494) distribution
F_crit <- qf(1 - alpha, df1 = q, df2 = df_ur)

# p-value: probability of observing F >= F_manual under H0
# lower.tail = FALSE gives the UPPER tail probability
p_val  <- pf(F_manual, df1 = q, df2 = df_ur, lower.tail = FALSE)

cat("Critical value (5%):", round(F_crit, 4), "\n")
cat("p-value:            ", round(p_val, 4), "\n")
```

Now make the decision --- exactly as described in Exercise 1, Step 5:

```{r decision-rule}
cat("F-statistic:", round(F_manual, 4), "\n")
cat("Critical value:", round(F_crit, 4), "\n\n")

if (F_manual > F_crit) {
  cat("DECISION: Reject H0 (F =", round(F_manual, 3),
      "> critical value =", round(F_crit, 3), ")\n")
  cat("Since H0 is TRUE in our DGP, this is a Type I error.\n")
  cat("This happens about 5% of the time --- exactly what alpha = 0.05 means.\n")
} else {
  cat("DECISION: Fail to reject H0 (F =", round(F_manual, 3),
      "<= critical value =", round(F_crit, 3), ")\n")
  cat("Correct decision --- H0 is indeed true in our DGP.\n")
  cat("The test correctly concluded that PrevExp does not affect wages.\n")
}
```

---

**Step 6. Verify against `linearHypothesis()`.**

The `car` package provides `linearHypothesis()`, which computes the $F$-statistic automatically from the unrestricted model. Your manual calculation from Steps 4--5 should match exactly.

```{r verify-names}
library(car)

# IMPORTANT: check what names R assigned to the coefficients
# The names you pass to linearHypothesis() must match these EXACTLY
names(coef(unrestricted))
```

```{r verify-test}
# Test H0: beta_3 = 0 AND beta_4 = 0
linearHypothesis(unrestricted, c("PrevExp = 0", "PrevExp2 = 0"))
```

\begin{bookexample}
\textbf{\color{blockcolor}Reading the \texttt{linearHypothesis()} output}

\medskip
The output shows a table with two rows:
\begin{itemize}
  \item \textbf{Row 1} (Model 1): the \emph{restricted} model. \texttt{Res.Df} $= 496$ is the residual df ($n - 3 - 1$). \texttt{RSS} $= SSR_r$.
  \item \textbf{Row 2} (Model 2): the \emph{unrestricted} model. \texttt{Res.Df} $= 494$ is the residual df ($n - 5 - 1$). \texttt{RSS} $= SSR_{ur}$.
\end{itemize}

\textbf{Check three things:}
\begin{enumerate}
  \item The \texttt{F} value matches your \texttt{F\_manual} from Step 4.
  \item The \texttt{Pr(>F)} value matches your \texttt{p\_val} from Step 5.
  \item The difference in \texttt{Res.Df} is $496 - 494 = 2 = q$, the number of restrictions.
\end{enumerate}

\medskip
\warn{Common mistake:} the variable names inside \texttt{c("PrevExp = 0", ...)} must \emph{exactly} match the names from \texttt{names(coef(unrestricted))}. If your data frame used different column names (e.g., \texttt{prev\_exp} instead of \texttt{PrevExp}), R would use those names instead, and you would need to match them. A typo produces an error, not a wrong answer --- so read the error message carefully.
\end{bookexample}

---

**Step 7 (take-home). What happens when $H_0$ is false?**

Go back to Step 1 and change the true parameters so that previous experience matters:

```r
b3 <- 0.015       # positive linear effect (was 0)
b4 <- -0.0003     # diminishing returns (was 0)
```

Then re-run Steps 2--6. You should observe:

- $SSR_r - SSR_{ur}$ is now **much larger** (dropping relevant variables hurts the fit substantially).
- The $F$-statistic is **large** (well above the critical value).
- The $p$-value is **tiny** ($p \ll 0.05$).
- The test **correctly rejects** $H_0$.

This is the **power** of the test: its ability to detect a false null. The larger the true effect ($\beta_3$, $\beta_4$), the larger $F$ becomes and the more likely we are to reject. Try different values of `b3` and `b4` to see how the $F$-statistic responds.

